Convergent Evolution During Local Adaptation to Patchy Landscapes

### 2014-07-03 patchy-selection-paper.tex

\title{
The Probability of Convergent Evolution During Local Adaptation. 
}

@@

\maketitle
\begin{center}
$^1$ Computational Biology and Bioinformatics \\ 
University of Southern California, Los Angeles, CA \\
$^2$Center for Population Biology \& Department of Evolution and Ecology \\ 
University of California -- Davis, Davis, CA, 95616
\end{center}

%%%%%%%%%%%
\abstract{
When the same phenotype is found to be responding to selection in parallel in isolated parts of a species range,
it provides strong evidence as to the degree of constraint imposed on evolution by the underlying genetic architecture.
Migration can counfound such inferences by synchronizing adaptation across the range;
but the importance of such sharing of local adaptations is unclear.
This paper studies the balance between these two in continuous environments with patchy selection pressures,
such as when an animal's coloration is selected to match its background,
but the species range contains large areas of differently colored substrate.
The main questions answered are: 
How long does it take for a new, locally adapted allele to appear in a patch of habitat through new mutation?
Or, through migration from another, already adapted patch?
Which is more likely to occur, as a function of patch separation?
How can we tell which has occurred?
These are answered by decomposing the migration--selection equilibrium
surrounding an already adapted patch
into families of migrant alleles,
in particular treating those rare families that reach new patches
as spatial branching processes.
The speed at which successful migrant alleles traverse regions where they are maladapted
determines the length of the haplotype shared by around shared, adaptive alleles,
which is compared to the length of the haplotype shared within a patch.
Cryptic coloration in the rock pocket mouse, \textit{Chaetodipus intermedius}, is discussed as an empirical example.

@@

The convergent evolution of similar phenotypes 
in response to shared selection pressures is a testiment to the power of selection 
to repeatedly sculpt phenotypic variation. 
In some cases this convergence extends to the molecular level, 
with disparate taxa converging to the same phenotype through parallel 
genetic changes in the same pathway, genes, 
or even by precisely the same genetic changes
\citep{stern2013genetic}. 
\plr{want to also cite \citep{zhen2012parallel,martin2013repeated}?}
Convergent adaptation can also occur within species, if individuals within a species 
adapt to the same environment, in parallel, through different genetics changes. 
There are a growing number of examples of this in a range of well studied organisms and phenotypes \cite{Arendt:08}.
Such evolution of convergent phenotypes is favored by higher mutational input, 
i.e.\ larger mutational target and/or population size \citep{softsweepsII}.
The geographic distribution of populations can also affect the probability of parallel mutation within a species:
a widespread species is more likely to adapt by multiple, parallel mutations if dispersal is geographically limited, 
since subpopulations will adapt via new mutations before the adaptive allele arrives via migration \citep{ralph2010parallel}. 
Standing variation for a trait can also increase the probability of convergence, 
as standing variation for an allele increases the probability that the selective sweep will be \emph{soft}
(beginning from a base of multiple copies),
leading to genetic patterns similar to convergent adaptation \citep{orr2001sieve,softsweepsI}.

@@

For the sake of simplicity, we assume a constant haploid population density $\rho$, 
across both types of habitat,
We further assume that the variance in offspring number is $\xi^2$, 
and that the mean squared distance between parent and child is $\sigma^2$
(i.e.\ $\sigma$ is the dispersal distance). Differences in population
densities on and off the patches may arise if selection is hard, in
that case these differences can be incorporated into the density \citep[see][]{lenormand2002limits}. 
We also aim to describe the scenario where it is sensible to think of large, discrete patches;
the case of a fine-scale heterogeneous environment is treated
\citep{elsewhere}. \gc{IS THIS LAST SENTENCE NEEDED IF SO GIVE CITATIOn}

@@

\gc{NOT SURE IF ANYONE HAS LOOKED AT THI IN LONG TIME?}
If we take the mutational influx in a patch to be one per 1,000 generations ($\rho A \mu = .001$)
and the population density to be 1 ($\rho =1$, so $\log(A\mu)\approx 7$)
then adaptation is mostly parallel for patches separated by gaps larger than $R/\sigma > (7+\log(s_m))/\sqrt{s_m}$.
If the selective pressure is fairly strong (say, $s_m=.05$),
then for migration to be slower than mutation,
the distance between patches must be at least 30 times the dispersal distance (i.e.\ $R/\sigma > 30$).
If $s_m$ is much smaller, say $s_m = .001$, 
then the patches must be separated by several hundred times the dispersal distance (i.e.\ $R/\sigma > 300$).
The situation does not change much if the influx is quite a bit higher -- even if $A \mu = 1$ (so $\log(A\mu) \approx 2.3$),
the required separation between patches is only reduced to $R/\sigma > 10$ with $s_m=.05$ and $R/\sigma>150$ for $s_m=.001$.
On the other hand, if the separation is much smaller than this transitional value, 
then adaptations are likely to be shared between patches.

@@

In this work, we have tried to emphasize an intuitive, individual-centered understanding of the process,
determining, for instance, 
the speed at which successful migrants traverse regions where they are maladapted,
and the effective rate at which new families of migrant alleles appear
at long distances from a patch.
These quantities may be as useful for applied work
as more abstract ones such as the probability that adaptation occurs by migration versus new mutation.

@@

\paragraph{Concluding thoughts}
The falling cost of population genomic sequencing means that we will soon have the
opportunity to study the interplay of adaptation with geography and ecology 
across many populations within a species. 
Our work suggests that even quite geographically close populations 
may be forced to adapt convergently when migration is geographically limited. 
Thus, systems where populations have been repeatedly subject to 
strong local selection pressures may offer the opportunity 
to study highly replicated convergent adaptation within a similar
genetic background \citep{stern2013genetic}.  
Such empricial work will also strongly inform our understanding of how gene flow 
may work to keep ecologically similar populations evolving in concert \citep{sexton2013genetic}.
Our results suggest that adaptation to shared environments is
certainly no guarantee of a shared genetic basis to adaptation, 
suggesting that rapid adaptation to a shared environment could potentially drive
speciation under some scenerios \citep[in a manner similar to ][]{Kondrashov:03}.

>>>>

\title{
Convergent Evolution During Local Adaptation to Patchy Landscapes
}

@@

\maketitle
\begin{center}

$^1$ Computational Biology and Bioinformatics \\ 
University of Southern California, Los Angeles, CA \\
$^2$Center for Population Biology \& Department of Evolution and Ecology \\ 
University of California -- Davis, Davis, CA, 95616
\end{center}

%%%%%%%%%%%
\abstract{
Populations within a species often repeatedly locally adapt to similar environmental conditions
across their range. Such adaptation can occur through convergent
evolution as different alleles arise and spread in different parts of
the range. Or it can occur through the spread of alleles by migration
acting to synchronizing adaptation across the species.
The tension between the two reflects the degree of 
constraint imposed on evolution by the underlying genetic
architecture versus how how effectively selection acts to inhibt the geographic spread of local
adapted alleles.
This paper studies a model of the the balance between these two routes to adaptation in continuous environments with patchy selection pressures.
We address the following questions: 
How long does it take for a new, locally adapted allele to appear in a
patch of favorable habitat through new mutation?
Or, through migration from another, already adapted patch?
Which is more likely to occur, as a function of patch separation?
How can we tell which has occurred, i.e. what population genetic
signal is left by the spread of migrant alleles and how long does this
signal persist for?
To answer these questions we decompose the migration--selection equilibrium
surrounding an already adapted patch
into families of migrant alleles,
in particular treating those rare families that reach new patches
as spatial branching processes.
This provides a set of tractable approximations to understand the role
of geographic separation between favorable patches in promoting
convergent adaptation and the genomic signals left.
We illustrate these ideas using the convergent evolution of cryptic coloration in the rock pocket mouse, \textit{Chaetodipus intermedius}, as an empirical example.

@@

The convergent evolution of similar phenotypes 
in response to shared selection pressures is a testiment to the power of selection 
to repeatedly sculpt phenotypic variation. 
In some cases this convergence extends to the molecular level, 
with disparate taxa converging to the same phenotype through parallel 
genetic changes in the same pathway, genes, 
or even by precisely the same genetic changes
\citep{stern2013genetic,zhen2012parallel,martin2013repeated}. 
Convergent adaptation can also occur within species, if individuals within a species 
adapt to the same environment, in parallel, through different genetics changes. 
There are a growing number of examples of this in a range of well studied organisms and phenotypes \cite{Arendt:08}.
Such evolution of convergent phenotypes is favored by higher mutational input, 
i.e.\ larger mutational target and/or population size \citep{softsweepsII}.
The geographic distribution of populations can also affect the probability of parallel mutation within a species:
a widespread species is more likely to adapt by multiple, parallel mutations if dispersal is geographically limited, 
since subpopulations will adapt via new mutations before the adaptive allele arrives via migration \citep{ralph2010parallel}. 
Standing variation for a trait can also increase the probability of convergence, 
as standing variation for an allele increases the probability that the selective sweep will be \emph{soft}
(beginning from a base of multiple copies),
leading to genetic patterns similar to convergent adaptation \citep{orr2001sieve,softsweepsI}.

@@

For the sake of simplicity, we assume a constant haploid population density $\rho$, 
across both types of habitat,
We further assume that the variance in offspring number is $\xi^2$, 
and that the mean squared distance between parent and child is $\sigma^2$
(i.e.\ $\sigma$ is the dispersal distance). Differences in population
densities on and off the patches may arise if selection is hard, in
that case these differences can be incorporated into the density \citep[see][]{lenormand2002limits}. 
%We also aim to describe the scenario where it is sensible to think of large, discrete patches;
%the case of a fine-scale heterogeneous environment is treated
%\citep{elsewhere}. \gc{IS THIS LAST SENTENCE NEEDED IF SO GIVE CITATIOn}

@@

\gc{NOT SURE IF ANYONE HAS LOOKED AT THIS NEXT PARAGRAPH IN LONG TIME?}
If we take the mutational influx in a patch to be one per 1,000 generations ($\rho A \mu = .001$)
and the population density to be 1 ($\rho =1$, so $\log(A\mu)\approx 7$)
then adaptation is mostly parallel for patches separated by gaps larger than $R/\sigma > (7+\log(s_m))/\sqrt{s_m}$.
If the selective pressure is fairly strong (say, $s_m=.05$),
then for migration to be slower than mutation,
the distance between patches must be at least 30 times the dispersal distance (i.e.\ $R/\sigma > 30$).
If $s_m$ is much smaller, say $s_m = .001$, 
then the patches must be separated by several hundred times the dispersal distance (i.e.\ $R/\sigma > 300$).
The situation does not change much if the influx is quite a bit higher -- even if $A \mu = 1$ (so $\log(A\mu) \approx 2.3$),
the required separation between patches is only reduced to $R/\sigma > 10$ with $s_m=.05$ and $R/\sigma>150$ for $s_m=.001$.
On the other hand, if the separation is much smaller than this transitional value, 
then adaptations are likely to be shared between patches.

@@

%commented out as wasn't quite sure what purpose this paragraph served.
% In this work, we have tried to emphasize an intuitive, individual-centered understanding of the process,
% determining, for instance, 
% the speed at which successful migrants traverse regions where they are maladapted,
% and the effective rate at which new families of migrant alleles appear
% at long distances from a patch.
% These quantities may be as useful for applied work
% as more abstract ones such as the probability that adaptation occurs by migration versus new mutation.

In developing the set of approximations in the paper we have ignored a
number of complicating factors. We now briefly discuss these.

@@

\paragraph{Concluding thoughts}
The falling cost of population genomic sequencing means that we will soon have the
opportunity to study the interplay of adaptation with geography and ecology 
across many populations within a species. 
Our work suggests that even quite geographically close populations 
may be forced to locally adapt convergently when migration is geographically limited. 
Thus, systems where populations have been repeatedly subject to 
strong local selection pressures may offer the opportunity 
to study highly replicated convergent adaptation within a similar
genetic background \citep{stern2013genetic}.  
Such empricial work will also strongly inform our understanding of the
ability of gene flow to keep ecologically similar populations evolving
in concert \citep{sexton2013genetic}.
Our results suggest that adaptation to shared environments is
certainly no guarantee of a shared genetic basis to adaptation, 
suggesting that rapid adaptation to a shared environment could potentially drive
speciation if the alleles that spread in each population fail to work
well together \citep{Kondrashov:03}.
